Towards the construction of a causal map of the human phenome

### 2017-06-26 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{\texttt{r\ format(Sys.time(),\ \textquotesingle{}\%d\ \%B\ \%Y\textquotesingle{})}}

@@

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is known as an instrumental variable
for the exposure, and can be exploited to mimic a randomised controlled
trial, making causal inference by comparing the outcome phenotypes
between those individuals that have the exposure-increasing allele
against those who do not. Multiple independent genetic variants for a
particular exposure can be used jointly to improve causal inference, the
premise being that each variant is an independent natural experiment,
and an overall causal estimate can be obtained by meta-analysing the
single estimates from each instrument.

@@

In this paper we introduce new machine learning approaches that attempt
to automate both instrument and method selection. Using curated GWAS
summary data for thousands of phenotypes, we use these new methods to
construct a graph of millions of causal estimates.

@@

@@

\textbf{Mean-based methods:} The inverse variance weighted (IVW)
meta-analysis approach assumes that variant exhibits no HP (fixed
effects meta-analysis) or that HP is present but balanced (random
effects meta-analysis). Egger regression relaxes the HP assumption
further by allowing the horizontal pleiotropy to systematically occur in
a specific direction, known as directional horizontal pleiotropy. The
Rucker framework uses estimates of heterogeneity to navigate between
these nested models. A jackknife approach (random selection with
replacement of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations.

The Rucker framework is a system used in meta-analysis to navigate
between IVW and Egger regression, and fixed and random effects models.
It was recently applied to MR to use heterogeneity statistics to select
the most appropriate model of pleiotropy, and therefore the most
appropriate MR method amongst the four mean-based estimators. There

@@

\textbf{Mode-based methods:} Supposing that

@@

\[
\begin{aligned}
\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y}) \times  \\
\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
\end{aligned}
\]

@@

>>>>

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{26 June 2017}

@@

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant is an
independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses.

@@

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to
automatically discard genetic variants that are likely to be invalid.
Second, we hypothesised that characteristics of the summary data could
indicate which method would be most reliable, and we introduce new
machine learning approaches that attempt to automate both instrument and
method selection. Using curated GWAS summary data for thousands of
phenotypes, we use these new methods to construct a graph of millions of
causal estimates. We consider this to be only a `first draft' of the
causal map of the human phenome.

@@

The use of summary data in two-sample MR is described in detail
elsewhere. A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained, collecting their effect sizes
and standard errors and effect alleles for the association with the
exposure trait. These can be obtanied manually from complete summary
data or from curated lists of significant GWAS associations. Next, the
effects of those SNPs on the outcome need to be obtained, typically
necessitating complete summary data because it is unlikely that these
SNPs will have reached genome-wide significance (and therefore be
present in curated catalogues). At its most simple implementation, the
regression of the SNP-exposure effect sizes against the SNP-outcome
effect sizes, with greater weight afforded to those SNPs with smaller
SNP-outcome standard errors, provides the estimate of the causal effect
of the exposure on the outcome.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there is sufficient summary data available.
Supplementary table X provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits). Supplementary table X provides a list of the
genetic instruments used for each trait.

@@

\textbf{Mean-based methods:} Here we consider four nested models. The
inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy be directional, where it systematically
occurs in a specific direction. Random effects Egger regression allows
heterogeneity around the directional HP, as long as the HP effects are
not correlated with the SNP-exposure effects (this is the INSIDE
assumption).

The Rucker framework uses estimates of heterogeneity in the IVW and
Egger frameworks to navigate between these nested models. A jackknife
approach (random selection with replacement of instruments) can be used
to obtain a sampling distribution for the model estimate amongst these
four variations. Using 1000 rounds of jackknife estimates, we can obtain
a final estimate using the mean or the median of the distribution. We
only use the jackknife approach for associations where there are 15 or
more variants to avoid combinatorial saturation.

@@

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments. This can be
extended in two ways, first by weighting the strength of each cluster by
the inverse of the variance of the SNP-outcome associations in the
cluster; second by weighting the strength of each cluster by the inverse
of the variance of the SNP-exposure associations.

@@

\[
\begin{aligned}
\beta_{gx,y,i*} & \sim N(\mu_{gx,y}, \sigma^2_{gx,y})  \\
\mu_{gx,y,i*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gx,y,i*} & \sim U(0.001, 0.01) \\
\beta_{gy,x,j*} & \sim N(\mu_{gy,x}, \sigma^2_{gy,x}) \\
\mu_{gy,x,j*} & \sim U(-0.005, 0.005) \\
\sigma^2_{gy,x,j*} & \sim U(0.001, 0.01) \\
\end{aligned}
\]

@@

\subsubsection{Graph database}\label{graph-database}

The set of MR estimates obtained from this analysis are recorded in a
Neo4j graph database. Because each association has up to 28 different
estimates, for simplicity we distill this down to a single `best
estimate' for each association using the following rules:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If the number of variants after Steiger filtering is greater than 5
  then apply the MoE to obtain the best method
\item
  If the number of variants after Steiger filtering is less than or
  equal to 5 but greater than 1 then use the IVW random effects approach
  on the filtered set of variants
\item
  If there is 1 variant retained in the Steiger approach then use the
  Wald ratio on the remaining variant
\item
  If there are no variants remaining after Steiger filtering then
  declare no causal association.
\end{enumerate}

For specific hypotheses we strongly recommend that estimates from all
sensitivity analyses are scrutinised and reported.

\subsection{Results}\label{results}

\subsubsection{Steiger filtering improves
reliability}\label{steiger-filtering-improves-reliability}

\subsubsection{Mixture of experts method selection improves over any
single
method}\label{mixture-of-experts-method-selection-improves-over-any-single-method}

Number of associations using different approaches Which methods were
preferentially chosen

\subsubsection{A first draft of the causal map of the human
phenome}\label{a-first-draft-of-the-causal-map-of-the-human-phenome}

How many associations reach nominal or FDR corrected significance How
many bi-directional associations How many associations could be declared
for a causal chain of 2

\subsection{Discussion}\label{discussion}

\begin{itemize}
\tightlist
\item
  Limitations of MR that cannot be addressed without biological
  information

  \begin{itemize}
  \tightlist
  \item
    Impossible causal associations
  \item
    Temporal effects
  \item
    Risk liability etc
  \item
    Single variants
  \end{itemize}
\item
  Further developments required to improve causal estimates
\item
  The use of causal chains to improve power
\end{itemize}
